---
title: 'SwitftRKey: A typing prediction app written in R'
author: "Marcio Gualtieri (marcio dot gualtieri at gmail dot com)"
output:
  html_document:
    toc: yes
    toc_depth: 6
  md_document:
    df_print: kable
    toc: yes
    toc_depth: 6
    variant: markdown_github
---

# `swiftrkey` R Package

## Overview

The purpose of this library is encapsulating the [text predicting model created previously](../notebook/) as well as code to generate predictions using the model.

## Installation Instructions

This package can be installed from GitHub using the `devtools` package. If you don't have it already, install `devtools` with the following command:

```{r, eval = FALSE}
install.packages("devtools")
```

Then run the following th install the text prediction library:

```{r, eval = FALSE}
library("devtools")
install_github("marciogualtieri/SwifRKey", subdir="R/model/swifrkey")
```

## Usage

First of all, load the library in the R environment:

```{r}
suppressMessages(library(swiftrkey))
```

You might get help by running the following command on RStudio's console:

```{r}
?suggestions
```

### Examples of Usage

```{r}
suggestions(model, "new", n = 3)
```

```{r}
suggestions(model, "new york", n = 3)
```

If you want to know the probabilities for each suggestion, you may run the following command instead:

```{r}
suggestions(model, "new", n = 3, with_probabilities = TRUE)
```

The `model` object is a Markov transition matrix, which is loaded with the library by default. In future releases I may add other models, e.g. for different languages (like`model_german` and `model_french`) or generated by different methods (I have used Katz backoff, for more details refer to the [notebook](../notebook)).


## Developer Guide

In the next sections, follow instructions for creating a R library for my own future reference.

### Setup

You will need the following packages for R package development:

```{r, eval = FALSE}
install.packages(c("devtools", "roxygen2", "knitr"))
install.packages("mixtools")
```

### Load the Libraries

```{r, eval = FALSE}
library("devtools")
library("roxygen2")
library("knitr")
```


### Create the Package Directory

```{r, eval = FALSE}
create("swiftrkey")
```

### Build & Reload Package

To build your package, go to RStudio's menu, then `Build` > `Build and Reload`, or using the command-line:

    $ R CMD INSTALL --no-multiarch --with-keep.source swiftrkey